Social inequality in mortality among adults and elderly in northern Sweden 1851--2013\footnote{'Socoeconomic inequalities and mortality I: Europe'. 41st Annual Meeting of the Social Science History Association, Chicago, November 17--20, 2016.}

### 2016-11-15 slides.Rnw
add stuff

\item Is social position \emp{equally important} among the \emp{retired} population as
  for those in \emp{working age}?
\end{enumerate}

@@

@@

\begin{itemize}
\item \emp{Swedish church book} records
\item digitized by the \emp{Demographic Data Base} at Umeå University, 
\begin{itemize}  
   \item \emp{\sc Popum} and 
   \item \emp{\sc Poplink}
\end{itemize}    
\item  \url{http://www.cedar.umu.se/}
\end{itemize}

@@

@@

\begin{itemize}
\item  \R : \emp{\url{https://www.R-project.org}}.
  \begin{itemize}
    \item \emp{RStudio}: \url{https://www.rstudio.com}.
  \end{itemize}    
\item \R\ \emp{packages} (\emp{\url{https://cran.R-project.org}}):
  \begin{itemize}
  \item \emp{skum} (Broström, 2016) data from \emp{DDB}.
  \item \emp{rmarkdown} (RStudio Team, 2016).
  \item \emp{knitr} (Xie, 2016).
  \item \emp{eha} (Broström, 2015).
  \item \emp{xtable} (Dahl, 2016).
\end{itemize}            
\end{itemize}

@@

\item[what] is this?

@@

\begin{enumerate}
\item \'Elite, {\sc Hisclass} \emp{1 and 2}: Higher managers and higher professionals.
\item Lower managers, {\sc Hisclass} \emp{3}.
\item Farmers and fishermen, {\sc Hisclass} \emp{8}.
\item Lower professionals, {\sc Hisclass} \emp{4, 5 and 6}: Lower professionals, lower clerical, foremen.
\item Medium skilled workers, {\sc Hisclass} \emp{7}.
\item Lower skilled workers, {\sc Hisclass} \emp{9, 10, 11 and 12}: Lower and unskilled workers (including farm workers)
%%\item Lower skilled farm workers, {\sc Hisclass} 10 and 12, Lower and unskilled farm workers
\end{enumerate}

@@

\begin{frame}{{\sc Poplink} data, 1901--1950}

<<getdata, results='asis'>>=
library(xtable)
library(knitr)
load("data/adult.rda")
adult <- cal.window(adult, c(1901, 1951))
source("R/addPer.R")
adult <- addPer(adult, cuts = seq(1901, 1951, by = 10))
adult <- rc(adult)
## NEW hisclass, 6 + 7 --> 6:
adult$hisclass[!is.na(adult$hisclass) & (adult$hisclass == 7)] <- 6
@

@@

\begin{frame}{{\sc Fob} data, 1961--2000}

<<ssykonfob>>=
load("fobb.rda")
@

@@

\begin{frame}{{\sc Lisa} data, 2002--2013}

<<lisadata>>=
load("liss.rda")
source("R/addPer.R")
liss$id <- liss$LINNEID
liss <- addPer(liss, name = "period", cuts = seq(2002, 2014, by = 6))
liss <- liss[!is.na(liss$hisclass), ]
liss$hisclass <- cut(liss$hisclass, c(0, 2.5, 4.5, 5.5, 8), 
                     labels = c("1+2", "3+4", "5", "6"))
liss <- liss[with(liss, !is.na(civst) &
                      !is.na(hisclass)), ]
liss$civst <- factor(liss$civst, 
                     labels = c("unmarried", "married", "prev. married"))
liss$urban <- liss$FORSAMLING %in% c(248001, 248002, 284004, 284005, 
                                     248011, 248201, 248204)
@

@@

\begin{frame}{Results, ages 40--59}

<<colorc>>=
## Colors for plotting strata:
scol <- c(rep("blue", 2), rep("green", 2), rep("red", 2))
## Common for all groups:
labb1 <- levels(adult$period)
labb2 <- levels(fobb$period)
labb <- c(labb1, labb2)
nper1 <- length(labb1)
nper2 <- length(labb2)
nper <- nper1 + nper2
fobb$urban <- fobb$FORSAMLING %in% c(248001, 248002, 284004, 284005, 248011, 248201, 248204)
civst <- numeric(NROW(fobb))
civst[fobb$CIVIL %in% c(0, 5)] <- 1 # Ogift
civst[fobb$CIVIL %in% c(2, 7)] <- 2 # Gift, sambo
civst[fobb$CIVIL %in% c(1, 3, 4, 6, 8, 9)] <- 3 # skild, anka
fobb$civst <- factor(civst, labels = c("unmarried", "married", "prev. married"))
source("R/getPH.R")
cap1w <-  "Cumulative hazards for HISCLASS by time period, ages 40--59, women."
scap1w <-  "Cumulative hazards, ages 40--59, women."
cap1m <-  "Cumulative hazards for HISCLASS by time period, ages 40--59, men."
scap1m <-  "Cumulative hazards, ages 40--59, men."
##
cap12 <- "Total hazard of dying before age 60 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom)."
scap12 <- "Total hazard by HISCLASS, ages 40--59."    
cap12log <- "Total hazard of dying before age 60 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom). Log scale."
scap12log <- "Total hazard by HISCLASS, ages 40--59 (log scale)."

cap2w <-  "Cumulative hazards for HISCLASS by time period, ages 60--89, women."
scap2w <-  "Cumulative hazards, ages 60--89, women."
cap2m <-  "Cumulative hazards for HISCLASS by time period, ages 60--89, men."
scap2m <-  "Cumulative hazards, ages 60--89, men."
##
cap22 <- "Total hazard of dying before age 90 for a 60 year old person by HISCLASS and decade, women (top) and men (bottom)."
scap22 <- "Total hazard by HISCLASS, ages 60--89."

##
adult4059 <- age.window(adult, c(40, 60))
fobb4059 <- age.window(fobb, c(40, 60))
## End 'common'
@

@@

\begin{frame}{Ages 40--59, women}

<<hazcalw>>=
resw <- getPH(adult4059[adult4059$sex == "female", ],
                fobb4059[fobb4059$KON == "kvinna", ],
                nper1, nper2, labb)
resm <- getPH(adult4059[adult4059$sex == "male", ],
                fobb4059[fobb4059$KON == "man", ],
                nper1, nper2, labb)
@

<<hazplotw, fig.height = 6>>=
source("R/plotHaz.ph.R")
plotHaz.ph(resw, main = labb, printLegend = FALSE)
@

@@

\begin{frame}{Ages 40--59, men}

<<hazplotm, fig.height = 6>>=
plotHaz.ph(resm, main = labb, printLegend = FALSE)
@

@@

\begin{frame}{Total hazards in ages 40--59, women}

<<cumsw, fig.height = 5.5>>=
source("R/plotED.ph.R")
woutp <- plotED.ph(resw, scol = scol, logScale = FALSE, ylab = "H(60|40)")
@

@@

\begin{frame}{Total hazards in ages 40--59, men}

<<cumsm, fig.height = 5.5>>=
moutp <- plotED.ph(resm, scol = scol, logScale = FALSE, ylab = "H(60|40)")
@

@@

\begin{frame}{Total hazards in ages 40--59, women: Log scale}

<<cumslogw>>=
wloutp <- plotED.ph(resw, scol = scol, logScale = TRUE, ylab = "H(60|40)",
main = "Women")
@

@@

\begin{frame}{Total hazards in ages 40--59, men: Log scale}
<<cumslogm>>=
mloutp <- plotED.ph(resm, scol = scol, logScale = TRUE, ylab = "H(60|40)",
main = "Men")
@

@@

\begin{frame}{{\sc Lisa} data, 2002--2013, women 40--59}

<<lissyoung>>=
liss4059 <- age.window(liss, c(40, 59))
fitw1 <- phreg(Surv(enter-40, exit-40, event) ~ strata(hisclass) + 
                   civst + urban, dist = "gompertz", 
               data = liss4059[liss4059$KON == "kvinna", ])
fitm1 <- phreg(Surv(enter-40, exit-40, event) ~ strata(hisclass) + 
                   civst + urban, dist = "gompertz", 
               data = liss4059[liss4059$KON == "man", ])
dcol <- c("blue", "darkgreen", "darkgreen", "red")
@

<<plotliss1>>=
plot(fitw1, fn = "cum", col = dcol, main = "Women", lwd = rep(2, 4))
@

@@

\begin{frame}{{\sc Lisa} data, 2002--2013, men 40--59}

<<plotliss1m>>=  
plot(fitm1, fn = "cum", col = dcol, main = "Men", lwd = rep(2, 4))
@

\begin{frame}{Cox regression, women 2002--2013, ages 40--59}

\scriptsize

<<tablisswom1, results = "asis">>=
fitwo1 <- coxreg(Surv(enter, exit, event) ~ hisclass + civst + urban + period, 
                 data = liss4059[liss4059$KON == "kvinna", ])
dr <- drop1(fitwo1, test = "Chisq")
ltx(fitwo1, dr = dr)
@

\begin{frame}{Cox regression, men 2002--2013, ages 40--59}

\scriptsize

<<tablissman1, results = "asis">>=
fitma1 <- coxreg(Surv(enter, exit, event) ~ hisclass + civst + urban + period, 
                 data = liss4059[liss4059$KON == "man", ])
dr <- drop1(fitma1, test = "Chisq")
ltx(fitma1, dr = dr)
@

\begin{frame}{Results, ages 60--89}

\begin{frame}{Ages 60--89, women}

<<getske6089>>=
adult6089 <- age.window(adult, c(60, 90))
fobb6089 <- age.window(fobb, c(60, 90))
resw <- getPH(adult6089[adult6089$sex == "female", ],
                fobb6089[fobb6089$KON == "kvinna", ],
                nper1, nper2, labb, from = 60)
resm <- getPH(adult6089[adult6089$sex == "male", ],
                fobb6089[fobb6089$KON == "man", ],
                nper1, nper2, labb, from = 60)
plotHaz.ph(resw, main = labb)

@

\begin{frame}{Ages 60--89, men}

<<hazplotmold>>=
plotHaz.ph(resm, main = labb)
@

\begin{frame}{Total hazards in ages 60--89, women}

<<cumswold>>=
woutp <- plotED.ph(resw, scol = scol, logScale = FALSE, ylab = "H(90|60)")  
@

\begin{frame}{Total hazards in ages 60--89, men}

<<cumsmold>>=
moutp <- plotED.ph(resm, scol = scol, logScale = FALSE, ylab = "H(90|60)")
@

\begin{frame}{{\sc Lisa} data, 2002--2013, women 60--89}

<<lissold>>=
liss6089 <- age.window(liss, c(60, 90))
fitw2 <- phreg(Surv(enter-60, exit-60, event) ~ strata(hisclass) + 
                   civst + urban, dist = "gompertz", 
               data = liss6089[liss6089$KON == "kvinna", ])
fitm2 <- phreg(Surv(enter-60, exit-60, event) ~ strata(hisclass) + 
                   civst + urban, dist = "gompertz", 
               data = liss6089[liss6089$KON == "man", ])
dcol <- c("blue", "darkgreen", "red", "red")
plot(fitw2, fn = "cum", col = dcol)

@

\begin{frame}{Cox regression, women 2002--2013, ages 60--89}

  \scriptsize
<<tablisswom2, results = "asis", eval = TRUE>>=
fitwo2 <- coxreg(Surv(enter, exit, event) ~ hisclass + civst + urban + period, 
                 data = liss6089[liss6089$KON == "kvinna", ])
dr <- drop1(fitwo2, test = "Chisq")
ltx(fitwo2, dr = dr)
@

\begin{frame}{{\sc Lisa} data, 2002--2013, men 60--89}

<<lissoldmen>>=
plot(fitm2, fn = "cum", col = dcol)
@

\begin{frame}{Cox regression, men 2002--2013, ages 60--89}

  \scriptsize
<<tablissman2, results = "asis", eval = TRUE>>=
fitma2 <- coxreg(Surv(enter, exit, event) ~ hisclass + civst + urban + period, 
                 data = liss6089[liss6089$KON == "man", ])
dr <- drop1(fitma2, test = "Chisq")
ltx(fitma2, dr = dr)
@

@@

%We started this paper by raising the issue of the development of social
%inequalities. When studying this question in history we often get surprised
%of its complexity. In a society characterised by large economic inequality, 
%we would expect that this would result in corresponding mortality
%differences, i.e. higher social classes having an advantage. We find the
%appearance of a consistent social pattern in mortality, to the advantage of
%the higher classes, but surprisingly late in our history. It is only in the
%recent decades that this pattern dominates. Improved survival characterised
%however all {\sc Hisclass}es. We can however confirm that there are gender
% differences in the social pattern. While there is no clear social gradient
% among men, the results at least partly indicate advantages for the higher
% social classes for women already early in the 20th century. Females had
% more to gain from a high social position. Previous studies of the mortality
% transition have made similar observations. Among adults, gender has often
% been of more significant importance than social class. The reason for this
% difference and for the late appearance of the modern social pattern among
% men is still an enigma that require further studies, even though we have
% suggested the possible importance of behavioural factors and changes in
% ideals connected to different social classes. This should stimulate us to
% further disentangle the pathways for how social position influence health
% and mortality in different societies. Finally, we do not find any clear
% differences in the social pattern in the two studied age groups.

>>>>

\item Is social position \emp{equally important} among the \emp{retired} population as
  for those in \emp{working age}?
\item Is there a difference between classes in mortality from
  \emp{cardiovascular} diseases and \emp{cancers} respectively? 
\item Are there large differences in survival depending on level of
  \emp{education} and \emp{income} categories?

\end{enumerate}

@@

\begin{itemize}
\item \emp{Town} founded in \emp{1622}
\begin{itemize}
\item  Substantially \emp{larger} than Skellefteå
\item \emp{administrative} centre of the county Västerbotten
\item \emp{Military} regiments.
\item \emp{Schools}
\item \emp{University} from \emp{1963--65}.  
\end{itemize}  
\item The rural part similar to the one in the Skellefteå region.
\end{itemize}

@@

\begin{itemize}
\item \emp{Swedish church book} records (1851--1950).
  \begin{itemize}
\item digitized by the \emp{Demographic Data Base} at Umeå University, 
\begin{itemize}  
   \item \emp{\sc Popum} and 
   \item \emp{\sc Poplink}
\end{itemize}    
\item  \url{http://www.cedar.umu.se/}
\end{itemize}
\item The \emp{Linnaeus} data base (1960--2013)
  \begin{itemize}
\item Census data 1960, 1965, \ldots, 1990
\item Population registers ({\sc Lisa}, 1986--2013)
  \begin{itemize}
    \item \emp{Income}
    \item \emp{Education}  
  \end{itemize}    
 \item Death information from \emp{National Board of Health and Welfare}
   \begin{itemize}
\item Death dates from 1 January 1961 to 31 December 2013.
  \item \emp{Causes of death} ({\sc Ulorsak}, ``main cause of death'')
   \end{itemize}
\end{itemize}
\end{itemize}

@@

\begin{frame}{The sampling frame}

<<lexis2>>=
plot(c(1851, 1851, 1951, 1951, 1851), c(40, 90, 90, 40, 40), 
     type = "l", xlim = c(1851, 2014), lwd = 2, col = "red",
     xlab = "Year", ylab = "Age", axes = FALSE, ylim = c(0, 90))
axis(1, at = c(1851, 1901, 1951, 1961, 1981, 2014), las = 2)
axis(2, las = 2, at = c(0, 40, 65, 90))
axis(4, las = 2, at = c(0, 40, 65, 90))
box()
abline(h = 0)
lines(c(1961, 1961, 2014, 2014, 1961), c(40, 90, 90, 40, 40), 
      col = "blue", lwd = 2)
text(1900, 65, "PopLink", cex = 2, col = "red")
text(1900, 55, "79593 persons")
text(1900, 50, "32801 deaths")
text(1985, 65, "Linnaeus", cex = 2, col = "blue")
text(1985, 55, "156676 persons")
text(1985, 50, "42879 deaths")
@ 
\end{frame}

\begin{frame}{The Linnaeus data}

\begin{columns}
\begin{column}{0.6\textwidth}
<<picklexus,fig.width=6,fig.height = 5.9>>=

load("data/vb.rda")
vbd <- vb[vb$event == 1 & vb$dataset != "poplink", c("birthdate", "exit", "sex")]
x <- vbd$birthdate + vbd$exit
y <- vbd$exit
##par(mfrow = c(1, 2))
plot(x[vbd$sex == "male"], y[vbd$sex == "male"], type = "p", 
     pch = ".", col = "blue", xlab = "Year", ylab = "Age", las = 2,
     axes = FALSE)
axis(1, at = c(seq(1960, 1990, by = 5), 2002, 2007, 2014), las = 2)
axis(2, at = c(40, 65, 90))
axis(4, at = c(40, 65, 90))
for (i in seq(1960, 1990, by = 5) + (365 - 61) / 365){
    abline(v = i, lwd = 2)
}
for (i in 2002:2013){
    abline(v = i, lwd = 2, col = "black")
}
abline(h = c(40, 65, 90), lty = 2)
points(x[vbd$sex == "female"], y[vbd$sex == "female"], type = "p", 
     pch = ".", col = "red")
box()
@

\end{column}

\begin{column}{0.4\textwidth}
\scriptsize

\vspace{3\baselineskip}

\begin{tabular}{l|rr}
      & Number & Deaths \\ \hline
\textcolor{red}{Women} & 70\,249 & \textcolor{red}{13\,888} \\
\emp{Men} & 86\,427 & \emp{28\,991} \\ \hline
Total  &156\,676 & 42\,879
\end{tabular}

\vspace{\baselineskip}

\emp{\underline{Note}}: Persons with \emp{missing {\sc Hisclass}} (occupation) are \emp{excluded}.
\end{column}
\end{columns}
\end{frame}

@@

\begin{itemize}
\item  \R : \emp{\url{https://www.R-project.org}}.
  \begin{itemize}
    \item \emp{RStudio}: \url{https://www.rstudio.com}.
  \end{itemize}    
\item \R\ \emp{packages} (\emp{\url{https://cran.R-project.org}}):
  \begin{itemize}
  \item \emp{skum} (Broström, 2016) data from \emp{DDB}.
  \item \emp{rmarkdown} (RStudio Team, 2016).
  \item \emp{knitr} (Xie, 2016).
  \item \emp{eha} (Broström, 2016).
  \item \emp{xtable} (Dahl, 2016).
\end{itemize}            
\end{itemize}

@@

\item[sex]
\item[birth date]
\item[death date]
\item[cause of death]
\item[parish]  
\item[\sc Hisclass] grouped.
\item[income] 1990--2005, grouped by quartiles.
\item[education] 1990--2005. Seven categories.

@@

\begin{enumerate}
\item \emp{\tt elite}, HISCLASS 1 and 2. Higher managers and professionals.
\item \emp{\tt lowMan}, HISCLASS 3 and 8. Lower managers and farmers.
\item \emp{\tt lowWhiteC}, HISCLASS 4, 5 and 6. Lower white collar.
\item \emp{\tt worker}, HISCLASS 7, 9, 10, 11 and 12. Workers of different skills including farm workers.
\end{enumerate}

@@

\begin{frame}{Proportional hazards for women?}

\includegraphics[width=11cm,height=8cm]{figure/hazplotw-1}

@@

\begin{frame}{Proportional hazards for men?}

\includegraphics[width=11cm,height=8cm]{figure/hazplotm-1}

@@

\begin{frame}[fragile]{Interactions}

\begin{verbatim}
fit <- coxph(Surv(enter - 40, exit - 40, event) ~ 
    period * hisclass * sex + urban, data = vb)
drop1(fit, test = "Chisq")
\end{verbatim}

\begin{center}
\begin{tabular}{lrrrr}
  \hline
 & Df & AIC & LRT & Pr($>$Chi) \\ 
  \hline
$<$none$>$ &  & 1631901.32 &  &  \\ 
  urban & 1 & 1631946 & 47.04 & $7.0\times 10^{-12}$ \\ 
  \emp{period:hisclass:sex} & 27 & 1631957 & 110.11 & \emp{$5.3\times 10^{-12}$} \\ 
   \hline
\end{tabular}
\end{center}

\begin{itemize}

\item This model is described by \emp{80 coefficients}.

\item We show them in \emp{graphs}.
\end{itemize}

@@

\begin{frame}{No proportional hazards (wrt \sc Hisclass)!}

\begin{itemize}
\item Fit each {\sc Hisclass} \emp{separately}, for each time period.
\item Use the \emp{cumulative hazard} at 65 (or 90) for comparison.
\end{itemize}

So, we fit separate PH models for each combination of

\begin{itemize}
  \item \emp{\tt sex}
  \item \emp{\tt period}: 1851--75, 1876--1900, 1901--10, etc.
  \item \emp{\tt hisclass}
  \item \emp{\tt age group}: 40--64, 65--89.
  \item \emp{\tt cause of death}
\end{itemize}    
\end{frame}

\begin{frame}{The data table}

<<lexis3, fig.height = 4>>=
plot(c(1851, 1851, 1951, 1951, 1851), c(40, 90, 90, 40, 40), 
     type = "l", xlim = c(1851, 2014), lwd = 2, col = "red",
     xlab = "Year", ylab = "Age", axes = FALSE, ylim = c(40, 90))
axis(1, at = c(1851, 1901, 1951, 1961, 1981, 2014), las = 2)
axis(2, las = 2, at = c(0, 40, 65, 90))
axis(4, las = 2, at = c(0, 40, 65, 90))
box()
abline(h = 0)
lines(c(1961, 1961, 2014, 2014, 1961), c(40, 90, 90, 40, 40), 
      col = "blue", lwd = 2)
for (i in c(1851, 1876, 1901, 1926, 1971, 1981, 1991, 2002, 2007)){
    lines(c(i, i), c(40, 90), lty = 2)
}
for (i in c(45, 50, 55, 60, 65, 70, 75, 80, 85)){
    lines(c(1851, 1951), c(i, i), lty = 2)
    lines(c(1961, 2014), c(i, i), lty = 2)
}
@

\emp{Poisson regression}:

In \emp{each cell}, and for \emp{each} combination of \emp{covariate level}s, calculate 
\begin{itemize}
\item  the number of \emp{deaths} and 
\item the total \emp{exposure time}.
\end{itemize}
\end{frame}

\begin{frame}{Women and men, age 40--64}

  \begin{center}
\includegraphics[height = 7.5cm]{figure/cumsw-1}
  \end{center}
\end{frame}

\begin{frame}{Relative differences, ages 40--64 and 65--89}

  %\begin{center}
\includegraphics[height = 7.5cm]{figure/cumswstand-1}\includegraphics[height=7.5cm]{figure/cumswold-1}
   %\end{center}
\end{frame}

\begin{frame}{Cancer mortality, ages 40-64 and 65-89}

\includegraphics[height=7.5cm]{figure/12cumsw-1}\includegraphics[height=7.5cm]{figure/12cumswold-1}

@@

\begin{frame}{Cardiovascular mortality}

\includegraphics[height=7.5cm]{figure/41cumsw-1}\includegraphics[height=7.5cm]{figure/41cumswold-1}

@@

\begin{frame}{Education and income, 1990--2005, all deaths}

\includegraphics[height=3.5cm]{figure/liswomfig-1} Women, 40--64.

\includegraphics[height=3.5cm]{figure/lismanfig-1} Men, 40--64.

@@

\begin{frame}{Education and income, 1990--2005, all deaths}

\includegraphics[height=3.5cm]{figure/liswomfig2-1} Women, 65--89.

\includegraphics[height=3.5cm]{figure/lismanfig2-1} Men, 65--89.

@@

\begin{frame}{Education and income, 1990--2005, cancer}

\includegraphics[height=3.5cm]{figure/12liswomfig-1} Women, 40--64.

\includegraphics[height=3.5cm]{figure/12lismanfig-1} Men, 40--64.

@@

\begin{frame}{Education and income, 1990--2005, cancer}

\includegraphics[height=3.5cm]{figure/12liswomfig2-1} Women, 65--89.

\includegraphics[height=3.5cm]{figure/12lismanfig2-1} Men, 65--89.

@@

\begin{frame}{Education and income, 1990--2005, cardiovascular}

\includegraphics[height=3.5cm]{figure/41liswomfig-1} Women, 40--64.

\includegraphics[height=3.5cm]{figure/41lismanfig-1} Men, 40--64.

@@

\begin{frame}{Education and income, 1990--2005, cardiovascular}

\includegraphics[height=3.5cm]{figure/41liswomfig2-1} Women, 65--89.

\includegraphics[height=3.5cm]{figure/41lismanfig2-1} Men, 65--89.

@@

%%% Stop here!

@@
